Causal Compatibility Inequalities Admitting of Quantum Violations in the Triangle Scenario

### 2017-01-27 documents/TriangleScenario/ts.tex
small typo
Added journals to references

% === All of the Packages I use frequently ===
\usepackage{../../packages/document_config}
\usepackage{../../packages/shared}
\usepackage{../../packages/misc_commands}
\usepackage{subcaption}
\usepackage{enumitem}

\renewcommand{\Events}[1]{\mathcal{S}\br{#1}}
\renewcommand{\tcdot}{\cdot}

@@

    \[ \probvec^{\mscenario} = M \tcdot \probvec^{\jointvar} \eq \label{eq:incidence_matrix_use} \]

@@

    Each of the above mentioned techniques suffers from computational complexity limitations. For example, the Fourier-Motzkin procedure is act worst case doubly exponential in the number of initial inequalities~\cite{Dantzig_1973}. For the purposes of this research, solving the marginal problem without reference to a marginal model was intractable. This will become apparent in \cref{sec:found_inequalities} when the inflation technique is applied to the triangle scenario, producing marginal models that are very large. Moreover, algorithms to partially solve the marginal problem such as Hardy-type hypergraph transversals were used to determine hundreds of thousands of causal compatibility inequalities for the triangle scenario; none of which were able to be violated quantumly using numerical optimizations. Therefore, simpler approaches to the marginal problem needed to be considered. \\

@@

    \begin{definition}
        \label{def:marginal_linear_program}
        \term{The Marginal Linear Program} is the following linear program:
        \begin{alignat*}{2}
            & \text{minimize:} \quad&& \emptyset \tcdot x\\
            & \text{subject to:} && x \succeq 0 \\
            & && M \tcdot x = \probvec^{\mscenario}
        \end{alignat*}
        If this ``optimization''\footnote{``Optimization'' is presented in quotes here because the minimization objective is trivially always zero ($\emptyset$ denotes the null vector of all zero entries). The primal value of the linear program is of no interest, all that matters is its \textit{feasibility}.} is \textit{feasible}, then there exists a vector $x$ than can satisfy \cref{eq:incidence_matrix_use} and is a valid joint distribution vector. Therefore feasibility implies that $\probvec^{\jointvar}$ exists and is precisely $x$. Moreover if the marginal linear program is \textit{infeasible}, then there \textit{does not} exist a joint distribution $\probvec^{\jointvar}$.
    \end{definition}

@@

    \begin{definition}
        \label{def:dual_marginal_linear_program}
        \term{The Dual Marginal Linear Program}:
        \begin{alignat*}{2}
            & \text{minimize:} \quad&& y \tcdot \probvec^{\mscenario}\\
            & \text{subject to:} && y \tcdot M \succeq 0
        \end{alignat*}
        Where $y$ is a real valued vector with the same length as $\probvec^{\mscenario}$.
    \end{definition}

    The dual marginal linear program not only answers the marginal problem for a specific marginal model $\probvec^{\mscenario}$ but as a by-product provides an inequality that witnesses its contextuality. If this is not obvious, first notice that the dual problem is \textit{never infeasible}; by choosing $y$ to be trivially the null vector $\emptyset$ of appropriate size, all constraints become satisfied. Secondly if the dual constraint $y \tcdot M \succeq 0$ holds \textit{and} the primal is feasible, then the following must hold:
    \[ y \tcdot \probvec^{\mscenario} =  y \tcdot M \tcdot x \geq 0 \eq \label{eq:dual_marginal_ineq} \]
    Therefore the \textit{sign} of the dual objective $d \defined \min \br{y \tcdot \probvec^{\mscenario}}$ classifies a marginal model's contextuality. If $d < 0$ then \cref{eq:dual_marginal_ineq} is violated and therefore $\probvec^{\mscenario}$ is contextual. Likewise if $d \geq 0$ (satisfying \cref{eq:dual_marginal_ineq}), then $\probvec^{\mscenario}$ is non-contextual.\footnote{Actually, if $d \geq 0$ then it is exactly $d = 0$ due to the existence of the trivial $y = \emptyset$. This observation is an instance of the \textit{Complementary Slackness Property} of~\cite{Bradley_1977}. Moreover, if $d < 0$, then it is unbounded $d = -\inf$. This latter point becomes clear upon recognizing that for any $y$ with $d < 0$, another $y' = \al y$ can be constructed (with $\al > 1$) such that $d' = \al d < d$. Since a more negative $d'$ can always be found, it must be that $d$ is unbounded. This is a demonstration of the fundamental \textit{Unboundedness Property} of~\cite{Bradley_1977}; if the dual is unbounded, then the primal is infeasible.} This is manifestation of Farkas's lemma~\cite{Schrijver_1998}. \\

    An \term{infeasibility certificate}~\cite{Andersen_2001} is any vector $y$ that satisfies the constraint of \cref{def:dual_marginal_linear_program} and also permits violation of \cref{eq:dual_marginal_ineq} for some marginal distribution vector $\probvec^{\mscenario}$.
    \[ y \in \R^{\abs{\Events{\mscenario}}} : y \tcdot M \succeq 0, \quad y \tcdot \probvec^{\mscenario} < 0 \]
    Most linear program software packages such as Mosek~\cite{Mosek_2016}, Gurobi~\cite{Gurobi_2016}, CPLEX~\cite{Cplex_2016} and CVX/CVX OPT~\cite{CVX_2016,CVX_Opt_2016} are capable of producing infeasibility certificates.
    Furthermore, for every $y$ satisfying $y \tcdot M \succeq 0$ there corresponds a \term{certificate inequality} that constraints the set of non-contextual marginal models. If $y$ is an infeasibility certificate, then all non-contextual marginal models $\probvec^{\mscenario}$ must satisfy \cref{eq:dual_marginal_ineq} making $y \cdot \probvec^{\mscenario} \geq 0$ a valid causal compatibility inequality.

@@

    \[ U = \bs{\prod_{m=1}^{d-1} \br{\prod_{n=m+1}^{d} \exp\br{i P_n \lambda_{n,m}}\exp\br{i \si_{m,n} \lambda_{m,n}}}} \tcdot \bs{\prod_{l=1}^{d} \exp\br{iP_l \lambda_{l,l}}}  \eq \label{eq:spengler_unitary} \]

@@

    \begin{definition}
        The \term{product distribution} of two disjoint probability distributions $\prob[V]$ and $\prob[W]$ (where $V \cap W = \emptyset$) is denoted as $\prob[V] \times \prob[W]$.
        \[ \forall s \in \Events{V \cup W} : \br{\prob[V] \times \prob[W]}\br{s} \defined \prob[V][s|_{V}]\tcdot\prob[W][s|_{W}] \]
        A product distribution of $k$ mutually disjoint distributions is defined as,
        \[ \prod_{i\in\bs{k}} \prob[V_i] \defined \br{\prob[V_1] \times \cdots \times \prob[V_k]} \]
    \end{definition}

@@

    If \cref{sec:unitary_group}, the SSH parameterization was introduced as a non-degenerate parameterization of the unitary group $\s{U}\br{d}$. As presented, the SSH parameterization suffers due to the presence of computationally expensive matrix exponential terms~\cite{Moler_2003}. Although not explicitly mentioned in~\cite{Spengler_2010_Unitary}, it is possible to remove the reliance on matrix exponential operations in \cref{eq:spengler_unitary} by utilizing the explicit form of the exponential terms in \cref{eq:exp_terms}. As a first step, recognize the defining property of the projective operators \cref{eq:projective_operator},
    \[ P_l^k = \br{\ket{l}\bra{l}}^k = \ket{l}\bra{l} = P_l \]
    This greatly simplifies the global phase terms $GP_l$,
    \[ GP_l = \exp\br{iP_l \lambda_{l,l}} = \sum_{k=0}^{\inf} \f{\br{iP_l \lambda_{l,l}}^k}{k!} = \ident + \sum_{k=1}^{\inf} \f{\br{i \lambda_{l,l}}^k}{k!}P_l^k = \ident + P_l \bs{\sum_{k=1}^{\inf} \f{\br{i \lambda_{l,l}}^k}{k!}} = \ident + P_l \br{e^{i \lambda_{l,l}} - 1} \eq \label{eq:unitary_GP} \]
    Analogously for the relative phase terms $RP_{n,m}$,
    \[ RP_{n,m} = \cdots = \ident + P_n \br{e^{i \lambda_{n,m}} - 1} \eq \label{eq:unitary_RP} \]
    Finally, the rotation terms $R_{m,n}$ can also be simplified by examining powers of $i \sigma_{n,m}$,
    \[ R_{m,n} = \exp\br{i \si_{m,n} \lambda_{m,n}} = \sum_{k=0}^{\inf} \f{\br{\ket{m}\bra{n} - \ket{n}\bra{m}}^k \lambda_{m,n}^k}{k!} \]
    One can verify that the following properties hold,
    \begin{align*}
        \br{\ket{m}\bra{n} - \ket{n}\bra{m}}^0 &= \ident \\
        \forall k \in \N, k \neq 0 : \br{\ket{m}\bra{n} - \ket{n}\bra{m}}^{2k} &= \br{-1}^k\br{\ket{m}\bra{m} + \ket{n}\bra{n}} \\
        \forall k \in \N : \br{\ket{m}\bra{n} - \ket{n}\bra{m}}^{2k+1} &= \br{-1}^k\br{\ket{m}\bra{n} - \ket{n}\bra{m}}
    \end{align*}
    Revealing the simplified form of $R_{m,n}$,
    \[ R_{m,n} = \ident + \br{\ket{m}\bra{m} + \ket{n}\bra{n}} \sum_{j=1}^{\inf} \br{-1}^j\f{\lambda_{n,m}^{2j}}{\br{2j}!} + \br{\ket{m}\bra{n} - \ket{n}\bra{m}} \sum_{j=0}^{\inf} \br{-1}^j\f{\lambda_{n,m}^{2j+1}}{\br{2j+1}!} \]
    \[ R_{m,n} = \ident + \br{\ket{m}\bra{m} + \ket{n}\bra{n}} \br{\cos\lambda_{n,m} - 1} + \br{\ket{m}\bra{n} - \ket{n}\bra{m}} \sin\lambda_{n,m} \eq \label{eq:unitary_R} \]
    By combining the optimizations of \cref{eq:unitary_GP,eq:unitary_RP,eq:unitary_R} together we arrive at an equivalent form for \cref{eq:spengler_unitary} that is computational more efficient.
    \[ U = \bs{\prod_{m=1}^{d-1} \br{\prod_{n=m+1}^{d} R_{m,n}RP_{n,m}}} \tcdot \bs{\prod_{l=1}^{d} GP_{l}}  \eq \label{eq:spengler_unitary_fast} \]

    \clearpage
    \section{Large Triangle Scenario Inequalities}
    \subsection{Web Inflation Inequality}
    \label{sec:web_inequality}
    The following is a causal compatibility inequality for the triangle scenario that is violated by the Fritz distribution outlined in \cref{sec:fritz_distribution} and was derived using the Web inflation of \cref{fig:the_web_inflation}.
    \begin{center}
        \includegraphics[width=\linewidth]{../../figures/inequalities/no_symwitness_ineq_print_standalone}
    \end{center}
    Note: $\prob\br{abc}$ is shorthand for $\prob[ABC]\br{abc}$.
    \clearpage
    \subsection{Symmetric Web Inflation Inequality}
    \label{sec:symmetic_web_inequality}
    The following is a symmetric causal compatibility inequality for the triangle scenario that is violated by the Fritz distribution outlined in \cref{sec:fritz_distribution} and was derived using the Web inflation of \cref{fig:the_web_inflation}.
    \begin{center}
        \includegraphics[width=\linewidth]{../../figures/inequalities/MosekCertFritzSym_ineq_standalone}
    \end{center}
    Note: $\prob\br{abc}$ is shorthand for $\prob[ABC]\br{abc}$ and $\bs{P(113)P(330)}_3$ is shorthand for the sum over permutations:
    \[ \bs{P(113)P(330)}_3 = P(113)P(330) + P(131)P(303) + P(311)P(033) \]

>>>>

% === All of the Packages I use frequently ===
\usepackage{../../packages/document_config}
\usepackage{../../packages/shared}
\usepackage{../../packages/misc_commands}

\renewcommand{\Events}[1]{\mathcal{S}\br{#1}} % This notation is conflicting

@@

    \[ \probvec^{\mscenario} = M \cdot \probvec^{\jointvar} \eq \label{eq:incidence_matrix_use} \]

@@

    Each of the above mentioned techniques suffers from computational complexity limitations. For example, the Fourier-Motzkin procedure is in the worst case doubly exponential in the number of initial inequalities~\cite{Dantzig_1973}. For the purposes of this research, solving the marginal problem without reference to a marginal model was intractable. This will become apparent in \cref{sec:found_inequalities} when the inflation technique is applied to the triangle scenario, producing marginal models that are very large. Moreover, algorithms to partially solve the marginal problem such as Hardy-type hypergraph transversals were used to determine hundreds of thousands of causal compatibility inequalities for the triangle scenario; none of which were able to be violated quantumly using numerical optimizations. Therefore, simpler approaches to the marginal problem needed to be considered. \\

@@

    \begin{definition}
        \label{def:marginal_linear_program}
        \term{The Marginal Linear Program} is the following linear program:
        \begin{alignat*}{2}
            & \text{minimize:} \quad&& \emptyset \cdot x\\
            & \text{subject to:} && x \succeq 0 \\
            & && M \cdot x = \probvec^{\mscenario}
        \end{alignat*}
        If this ``optimization''\footnote{``Optimization'' is presented in quotes here because the minimization objective is trivially always zero ($\emptyset$ denotes the null vector of all zero entries). The primal value of the linear program is of no interest, all that matters is its \textit{feasibility}.} is \textit{feasible}, then there exists a vector $x$ than can satisfy \cref{eq:incidence_matrix_use} and is a valid joint distribution vector. Therefore feasibility implies that $\probvec^{\jointvar}$ exists and is precisely $x$. Moreover if the marginal linear program is \textit{infeasible}, then there \textit{does not} exist a joint distribution $\probvec^{\jointvar}$.
    \end{definition}

@@

    \begin{definition}
        \label{def:dual_marginal_linear_program}
        \term{The Dual Marginal Linear Program}:
        \begin{alignat*}{2}
            & \text{minimize:} \quad&& y \cdot \probvec^{\mscenario}\\
            & \text{subject to:} && y \cdot M \succeq 0
        \end{alignat*}
        Where $y$ is a real valued vector with the same length as $\probvec^{\mscenario}$.
    \end{definition}

    The dual marginal linear program not only answers the marginal problem for a specific marginal model $\probvec^{\mscenario}$ but as a by-product provides an inequality that witnesses its contextuality. If this is not obvious, first notice that the dual problem is \textit{never infeasible}; by choosing $y$ to be trivially the null vector $\emptyset$ of appropriate size, all constraints become satisfied. Secondly if the dual constraint $y \cdot M \succeq 0$ holds \textit{and} the primal is feasible, then the following must hold:
    \[ y \cdot \probvec^{\mscenario} =  y \cdot M \cdot x \geq 0 \eq \label{eq:dual_marginal_ineq} \]
    Therefore the \textit{sign} of the dual objective $d \defined \min \br{y \cdot \probvec^{\mscenario}}$ classifies a marginal model's contextuality. If $d < 0$ then \cref{eq:dual_marginal_ineq} is violated and therefore $\probvec^{\mscenario}$ is contextual. Likewise if $d \geq 0$ (satisfying \cref{eq:dual_marginal_ineq}), then $\probvec^{\mscenario}$ is non-contextual.\footnote{Actually, if $d \geq 0$ then it is exactly $d = 0$ due to the existence of the trivial $y = \emptyset$. This observation is an instance of the \textit{Complementary Slackness Property} of~\cite{Bradley_1977}. Moreover, if $d < 0$, then it is unbounded $d = -\inf$. This latter point becomes clear upon recognizing that for any $y$ with $d < 0$, another $y' = \al y$ can be constructed (with $\al > 1$) such that $d' = \al d < d$. Since a more negative $d'$ can always be found, it must be that $d$ is unbounded. This is a demonstration of the fundamental \textit{Unboundedness Property} of~\cite{Bradley_1977}; if the dual is unbounded, then the primal is infeasible.} This is manifestation of Farkas's lemma~\cite{Schrijver_1998}. \\

    An \term{infeasibility certificate}~\cite{Andersen_2001} is any vector $y$ that satisfies the constraint of \cref{def:dual_marginal_linear_program} and also permits violation of \cref{eq:dual_marginal_ineq} for some marginal distribution vector $\probvec^{\mscenario}$.
    \[ y \in \R^{\abs{\Events{\mscenario}}} : y \cdot M \succeq 0, \quad y \cdot \probvec^{\mscenario} < 0 \]
    Most linear program software packages such as Mosek~\cite{Mosek_2016}, Gurobi~\cite{Gurobi_2016}, CPLEX~\cite{Cplex_2016} and CVX/CVX OPT~\cite{CVX_2016,CVX_Opt_2016} are capable of producing infeasibility certificates.
    Furthermore, for every $y$ satisfying $y \cdot M \succeq 0$ there corresponds a \term{certificate inequality} that constraints the set of non-contextual marginal models. If $y$ is an infeasibility certificate, then all non-contextual marginal models $\probvec^{\mscenario}$ must satisfy \cref{eq:dual_marginal_ineq} making $y \cdot \probvec^{\mscenario} \geq 0$ a valid causal compatibility inequality.

@@

    \[ U = \bs{\prod_{m=1}^{d-1} \br{\prod_{n=m+1}^{d} \exp\br{i P_n \lambda_{n,m}}\exp\br{i \si_{m,n} \lambda_{m,n}}}} \cdot \bs{\prod_{l=1}^{d} \exp\br{iP_l \lambda_{l,l}}}  \eq \label{eq:spengler_unitary} \]

@@

    \begin{definition}
        The \term{product distribution} of two disjoint probability distributions $\prob[V]$ and $\prob[W]$ (where $V \cap W = \emptyset$) is denoted as $\prob[V] \times \prob[W]$.
        \[ \forall s \in \Events{V \cup W} : \br{\prob[V] \times \prob[W]}\br{s} \defined \prob[V][s|_{V}]\cdot\prob[W][s|_{W}] \]
        A product distribution of $k$ mutually disjoint distributions is defined as,
        \[ \prod_{i\in\bs{k}} \prob[V_i] \defined \br{\prob[V_1] \times \cdots \times \prob[V_k]} \]
    \end{definition}

@@

    If \cref{sec:unitary_group}, the SSH parameterization was introduced as a non-degenerate parameterization of the unitary group $\s{U}\br{d}$. As presented, the SSH parameterization suffers due to the presence of computationally expensive matrix exponential terms~\cite{Moler_2003}. Although not explicitly mentioned in~\cite{Spengler_2010_Unitary}, it is possible to remove the reliance on matrix exponential operations in \cref{eq:spengler_unitary} by utilizing the explicit form of the exponential terms in \cref{eq:exp_terms}. As a first step, recognize the defining property of the projective operators \cref{eq:projective_operator},
    \[ P_l^k = \br{\ket{l}\bra{l}}^k = \ket{l}\bra{l} = P_l \]
    This greatly simplifies the global phase terms $GP_l$,
    \[ GP_l = \exp\br{iP_l \lambda_{l,l}} = \sum_{k=0}^{\inf} \f{\br{iP_l \lambda_{l,l}}^k}{k!} = \ident + \sum_{k=1}^{\inf} \f{\br{i \lambda_{l,l}}^k}{k!}P_l^k = \ident + P_l \bs{\sum_{k=1}^{\inf} \f{\br{i \lambda_{l,l}}^k}{k!}} = \ident + P_l \br{e^{i \lambda_{l,l}} - 1} \eq \label{eq:unitary_GP} \]
    Analogously for the relative phase terms $RP_{n,m}$,
    \[ RP_{n,m} = \cdots = \ident + P_n \br{e^{i \lambda_{n,m}} - 1} \eq \label{eq:unitary_RP} \]
    Finally, the rotation terms $R_{m,n}$ can also be simplified by examining powers of $i \sigma_{n,m}$,
    \[ R_{m,n} = \exp\br{i \si_{m,n} \lambda_{m,n}} = \sum_{k=0}^{\inf} \f{\br{\ket{m}\bra{n} - \ket{n}\bra{m}}^k \lambda_{m,n}^k}{k!} \]
    One can verify that the following properties hold,
    \begin{align*}
        \br{\ket{m}\bra{n} - \ket{n}\bra{m}}^0 &= \ident \\
        \forall k \in \N, k \neq 0 : \br{\ket{m}\bra{n} - \ket{n}\bra{m}}^{2k} &= \br{-1}^k\br{\ket{m}\bra{m} + \ket{n}\bra{n}} \\
        \forall k \in \N : \br{\ket{m}\bra{n} - \ket{n}\bra{m}}^{2k+1} &= \br{-1}^k\br{\ket{m}\bra{n} - \ket{n}\bra{m}}
    \end{align*}
    Revealing the simplified form of $R_{m,n}$,
    \[ R_{m,n} = \ident + \br{\ket{m}\bra{m} + \ket{n}\bra{n}} \sum_{j=1}^{\inf} \br{-1}^j\f{\lambda_{n,m}^{2j}}{\br{2j}!} + \br{\ket{m}\bra{n} - \ket{n}\bra{m}} \sum_{j=0}^{\inf} \br{-1}^j\f{\lambda_{n,m}^{2j+1}}{\br{2j+1}!} \]
    \[ R_{m,n} = \ident + \br{\ket{m}\bra{m} + \ket{n}\bra{n}} \br{\cos\lambda_{n,m} - 1} + \br{\ket{m}\bra{n} - \ket{n}\bra{m}} \sin\lambda_{n,m} \eq \label{eq:unitary_R} \]
    By combining the optimizations of \cref{eq:unitary_GP,eq:unitary_RP,eq:unitary_R} together we arrive at an equivalent form for \cref{eq:spengler_unitary} that is computational more efficient.
    \[ U = \bs{\prod_{m=1}^{d-1} \br{\prod_{n=m+1}^{d} R_{m,n}RP_{n,m}}} \cdot \bs{\prod_{l=1}^{d} GP_{l}}  \eq \label{eq:spengler_unitary_fast} \]

    \clearpage
    \section{Large Triangle Scenario Inequalities}
    \subsection{Web Inflation Inequality}
    \label{sec:web_inequality}
    The following is a causal compatibility inequality for the triangle scenario that is violated by the Fritz distribution outlined in \cref{sec:fritz_distribution} and was derived using the Web inflation of \cref{fig:the_web_inflation}.
    \begin{center}
        % TODO [TC] replace this inequality with correct one
        \includegraphics[width=\linewidth]{../../figures/inequalities/no_symwitness_ineq_print_standalone}
    \end{center}
    Note: $\prob\br{abc}$ is shorthand for $\prob[ABC]\br{abc}$.
    \clearpage
    \subsection{Symmetric Web Inflation Inequality}
    \label{sec:symmetic_web_inequality}
    The following is a symmetric causal compatibility inequality for the triangle scenario that is violated by the Fritz distribution outlined in \cref{sec:fritz_distribution} and was derived using the Web inflation of \cref{fig:the_web_inflation}.
    \begin{center}
        % TODO [TC] replace this inequality with correct one
        \includegraphics[width=\linewidth]{../../figures/inequalities/MosekCertFritzSym_ineq_standalone}
    \end{center}
    Note: $\prob\br{abc}$ is shorthand for $\prob[ABC]\br{abc}$ and $\bs{P(113)P(330)}_3$ is shorthand for the sum over permutations:
    \[ \bs{P(113)P(330)}_3 = P(113)P(330) + P(131)P(303) + P(311)P(033) \]
